Estimating Comparative Advantage in Improved Seed Adoption: Hybrid Maize in Ethiopia

### 2021-12-14 paper.tex
Updates from Overleaf

\usepackage{natbib}
\bibliographystyle{rusnat}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{amsmath}

@@

\begin{itemize}
    \item Tim Weiss- hybrid subsidies
    \item is there such a thing as traditional seeds?
    \item Let's evaluate the effect of adoption in the context of Ethiopia
    \item find ambiguous results, Why?
    \item have access to DNA fingerprinting data
    \item what do they say about adoption?
    \item turns out that "adoption" isn't really a reliable concept/variable
    \item most farmers are "adopting" under pretty conservative purity definitions
    \item Let's see what happens with different varieties of seed, does adoption of that seed lead to positive yield effects?
    \item can't use these methods but newer data has access to newer varieties where we're relatively more confident
\end{itemize}

@@

Food security is a major challenge in many East African countries-- Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, as of today, it constitutes the most widespread crop in the country.\footnote{FAO stats data shows the total area cultivated with Maize in 2019 is 25\% larger than the one cultivated with the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and Barley, respectively.} As Figure \ref{fig:maize_yields} shows, there is an upward trend in the total area cultivated by maize, with average yields that have almost double during the period.

@@

@@

In our paper, we build on that literature by investigating hybrid maize seed adoption in Ethiopia, focusing on how heterogeneous returns are also driven by climate and how this can affect dis-adoption decisions. We extend the role of heterogeneity in a household’s decision to adopt new technologies in four fronts.

@@

There were significant differences between the analysis carried out in ESS 1-3 and the most recent, ESS4. Most notably, the ESS4 is no longer a panel dataset made up of the households of ESS 1-3, but it includes important information on the DNA fingerprinting of seed type and misclassification rates. This is important information as the first three waves of ESS only included self-reported hybrid maize status. Since our estimation strategy requires a balanced panel dataset of maize growers, our main analysis will include the first three waves of the ESS. We use the fourth wave later for robustness and validation.

@@

\input{results/tables/summary}

@@

\begin{figure}
    \centering
    \includegraphics{results/figures/adoption_r4.png}
    \caption{Adoption rates of improved varieties in 2018 (by self-report, purity level, crop type, source and year of release)}
    \figcaption{Note: Shares represent the proportion of adopters among maize farmers. Acronyms include: SR1 - Self reported 1 [=1 if New, 2nd gen or recycled, =0 if Traditional]; SR2 - Self reported 1 [=1 if New, 2nd gen or recycled, =0 if Traditional]; DNA70 - Improved if purity $>=$70$\%$; DNA90 - Improved if purity $>=$90$\%$; DNA95 - Improved if purity $>=$95$\%$; HYB - Improved DNA, [=1 Hybrid, 0=Open pollinated]; DTM - Drought tolerant Maize [=1 DTM, 0=otherwise], CGIAR source [=1 if germplasm associated to a CGIAR-related thread], Exotic source [=1 if germplasm associated to an exotic source], Released 2000+ [=1 if DNA associated to a variety released on year 2000 or later], Released 2010+ [=1 if DNA associated to a variety released on year 2010 or later]. QPM is not included as there are too few households planting QPM seeds (n=6).}
    \label{fig:adoption_r4}
\end{figure}

@@

The advantage of our estimation strategy is that we do not rely on identification from an instrument or the structure of a model, but from the fact that the trajectory that a household takes in their adoption decisions through time is an informative source of identifying information. See \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} for more information on the assumptions involved for the Group Random Coefficients estimator.

@@

Implications

@@

\input{results/tables/switch1}
\input{results/tables/switch2}
\input{results/tables/switch3}

>>>>

\usepackage{natbib}
\bibliographystyle{rusnat}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{float}

@@

%%%% I added the bullet points where I think they fit what we have as an introduction

\begin{itemize}
    \item Tim Weiss- hybrid subsidies
    \item is there such a thing as traditional seeds?

    \item Let's evaluate the effect of adoption in the context of Ethiopia
    \item find ambiguous results, Why?
    \item have access to DNA fingerprinting data
    \item what do they say about adoption?
    \item turns out that "adoption" isn't really a reliable concept/variable
    \item most farmers are "adopting" under pretty conservative purity definitions
    \item Let's see what happens with different varieties of seed, does adoption of that seed lead to positive yield effects?
    \item can't use these methods but newer data has access to newer varieties where we're relatively more confident
\end{itemize}

@@

%%% Why it is important
Food security is a major challenge in many East African countries-- Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, as of today, it constitutes the most widespread crop in the country.\footnote{FAO stats data shows the total area cultivated with Maize in 2019 is 25\% larger than the one cultivated with the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and Barley, respectively.} As Figure \ref{fig:maize_yields} shows, there is an upward trend in the total area cultivated by maize, with average yields that have almost double during the period.

@@

%%%% Let's evaluate the effect of adoption in the context of Ethiopia

@@

%% Expand a bit on the methods
In our paper, we build on that literature by investigating hybrid maize seed adoption in Ethiopia, focusing on how heterogeneous returns XXXXXXXXXXXXXXXX

%% What we find

Our results show evidence that showcase that the average yields without adoption for

return to adoption
% Evidence of heterogeneity in response to hybrid adoption for joiners and leavers, however their average yield without adoption ($\mu_{01}$,$\mu_{10}$) are very similar.\\

% are also driven by climate and how this can affect dis-adoption decisions. We extend the role of heterogeneity in a household’s
% decision to adopt new technologies in four fronts.

@@

There were significant differences between the analysis carried out in ESS 1-3 and the most recent, ESS4. Most notably, the ESS4 is no longer a panel dataset made up of the households of ESS 1-3, but it includes important information on the DNA fingerprinting of seed type, with which we can calculate misclassification rates. This is important information as the first three waves of ESS only included self-reported hybrid maize status. Since our estimation strategy requires a balanced panel dataset of maize growers, our main analysis will include the first three waves of the ESS. We use the fourth wave later for validation, and to explore the impact of specific types of improved seeds.

@@

% \resizebox{0.7\textwidth}{!}{
{\small\tabcolsep=3pt  % hold it local
    \input{results/tables/summary}
}

@@

\subsection{Improved seeds}

Defining what constitutes an improved seed is challenging because of the different characteristics of seeds. A crop seed is considered “improved” in Ethiopia if it was tested by breeders and evaluated to be superior to existing varieties (MoA, 2013, SPIA, 2020). In Ethiopia, the government controls most of the seed system. Local agricultural offices or extension agents assess the demands of different varieties, the MoA decides the seed quantities to be produced, and actors such as regional seed enterprises, seed companies, seed unions, are involved in the production of certified seeds; and seeds are primarily distributed by seed producer cooperatives (SPCs), which account for 37\% of total distribution in the country (SPIA, 2020).

An improved seed can be obtained directly from such distribution centers (which would distribute a first generation seed), or it could be cultivated from second or posterior generations from the originally bred seed. Farmers are asked to report whether the variety they used was a first-generation improved seed, second generation,  recycled, or a traditional variety. When buying seeds, farmers might be able to identify the variety they are looking for, but it is unlikely that they would be aware of the source.

Fifty four varieties have been released in Ethiopia since 1990, out of which seven were released between 1990 to 1999, 29 between 2000 to 2009, and 18 between 2010 to 2019. Thirty four of such varieties contain CIMMYT germplasm, out of which, 10 Drought-tolerant maize varieties (DTMZ) and 8 Quality protein maize (QPM) were released (most of these between 2010 to 2019). Out of the CIMMYT-related germplasm, ten are Open-pollinated varieties (OPVs) and 25 are hybrid. 
The fourth wave of the ESS introduced for the first time DNA fingerprinting for barley, maize, and sorghum. The collection of crop-cuts for crop yield estimation facilitated the process for DNA fingerprinting. For the crop-cuts, a 4 square meter quadrant was randomly selected in each plot, so that total production harvested from such patch was weighted (both when freshly harvested and when dried). The DNA material was extracted from samples from such dried crop-cuts, that were then matched with the genetic material of the maize DNA reference library for Ethiopia, obtained from a CIMMYT and EIAR project. The collection of DNA samples focused only on eight regions. In each EA of such regions, a random sample of plots was selected, of a maximum of 10 fields for each crop. As a result, 505 plots were selected for DNA fingerprinting, which corresponded to 447 households.

The DNA fingerprinting process matched the genetic material of each sampled harvest with that of the reference library. The resulting matching process returns the name of the crop variety matched, its germplasm origin or their pedigree (whether they are a CIMMYT line, EIAR, IITA or private), the breeder or maintainer, and the year of their release (which could be indicative of the number of generations a seed had been cultivated), as well as the genetic purity of the sample (the degree of no contamination by other genetic varieties). It also informs whether the harvested seeds are open-pollinated varieties (OPVs) or hybrids; drought-tolerant maize varieties (DMTZ) or quality protein maize (QPM). This rich information about seed varieties provides a unique opportunity to further understand impacts differentiated by type of seed, degree of purity, year of release or source. Figure \ref{fig:adoption_r4} shows adoption rates in the fourth round for improved varieties, using the above characteristics as different definitions.

We find that 38\% of households self-report adopting a “new” improved variety (SR2 in Figure \ref{fig:adoption_r4}), and 43\% report adopting a “new”, “second generation” or “recycled” improved variety (SR1). Considering a household as adopter if the household planted an improved variety in at least one of their plots, all households where a DNA sample was obtained used a maize improved seed, considering a purity threshold of 70\% (DNA 70). When purity thresholds of 90 and 95\% are considered (DNA 90, DNA 95 in Figure \ref{fig:adoption_r4}, respectively), adoption rates change to 91 and 54\% of households, respectively. 62\% of households adopted a High-yielding variety, compared to 38\% that adopted an open-pollinated variety, and 15\% of households adopted a Drought-tolerant maize variety. 63\% of households that adopted an improved variety are using a CGIAR-derived germplasm, and 79\% are using an exotic germplasm. Finally, 64\% of households adopted a variety that was released on the year 2000 or after, and 24\% adopted a variety that was adopted in the year 2010 or after.

\begin{figure}
    \centering
    \includegraphics{results/figures/adoption_r4.png}
    \caption{Adoption rates of improved varieties in 2018 (by self-report, purity level, crop type, source and year of release)}
    \figcaption{Note: Shares represent the proportion of adopters among maize farmers. Acronyms include: SR1 [=1 if New, 2nd gen or recycled, =0 if Traditional]; SR2  [=1 if New, =0 if 2nd gen, recycled or traditional]; DNA70 - Improved if purity $>=$70$\%$; DNA90 - Improved if purity $>=$90$\%$; DNA95 - Improved if purity $>=$95$\%$; HYB - Improved DNA, [=1 Hybrid, 0=Open pollinated]; DTM - Drought tolerant Maize [=1 DTM, 0=otherwise], CGIAR source [=1 if germplasm associated to a CGIAR-related thread], Exotic source [=1 if germplasm associated to an exotic source], Released 2000+ [=1 if DNA associated to a variety released on year 2000 or later], Released 2010+ [=1 if DNA associated to a variety released on year 2010 or later]. QPM is not included as there are too few households planting QPM seeds (n=6).}
    \label{fig:adoption_r4}
\end{figure}

These ratios are in line with prior findings. In the regions of Oromia, Amhara, Tigray, and SNNPR, adoption rates of improved maize varieties of 39\% of households were found, according to self-reported information (Zeng et al., 2015). In the regions of Amhara, Benishangul-Gumuz, Oromia, SNNPR, and Tigray, adoption rates of improved maize varieties were found for 27\% of households (Jaleta et al., 2018). In Oromia, SNNPR, and Benishangul-Gumuz adoption of improved varieties per farmers’ self-reports was found for 84\% of maize growers in 2011, and 88\% of maize growers in 2013 (Yirga et al., 2017).

Understanding the effects of improved seed adoption is not only challenging because there could be multiple ways of establishing what constitutes an improved seed, but also because such definitions may or may note overlap, or because farmers may also missclassify a variety as improved. Figure \ref{fig:missclassification} shows the correlation between the different definitions discussed above. For those seeds for which genetic material was obtained, the correlation with the self-reported seeds indicator is low, especially when using lower cutoff thresholds.

\begin{figure}
    \centering
    \includegraphics{results/figures/missclassification.png}
    \caption{Correlation of different definitions of improved seeds, 4th round}
    \figcaption{Note: Shares represent the proportion of adopters among maize farmers. Acronyms include: SR1 [=1 if New, 2nd gen or recycled, =0 if Traditional]; SR2  [=1 if New, =0 if 2nd gen, recycled or traditional]; DNA70 - Improved if purity $>=$70$\%$; DNA90 - Improved if purity $>=$90$\%$; DNA95 - Improved if purity $>=$95$\%$; HYB - Improved DNA, [=1 Hybrid, 0=Open pollinated]; DTM - Drought tolerant Maize [=1 DTM, 0=otherwise], CGIAR source [=1 if germplasm associated to a CGIAR-related thread], Exotic source [=1 if germplasm associated to an exotic source], Released 2000+ [=1 if DNA associated to a variety released on year 2000 or later], Released 2010+ [=1 if DNA associated to a variety released on year 2010 or later]. QPM is not included as there are too few households planting QPM seeds (n=6).}
    \label{fig:missclassification}
\end{figure}

@@

\subsection{Group Random Coefficient Model}

The advantage of our estimation strategy is that we do not rely on identification from a valid instrument, but modelling the heterogeneity directly through a projection of their returns to adoption on their observed "trajectory" of adoption. This allows us to directly capture two effects: the heterogeneous effects of the so-called "comparative advantage" of adoption, and, under the assumption of linearity in comparative advantage, a description of the overall economy's sorting of farmers based on their comparative advantage. The first is important because it is able to disentangle the effect of a general household's agricultural ability (which can be thought of as a fixed effect) and their adoption-specific potential. The sorting parameter can tell us whether the economy sorts such that those that would benefit the most adoption actually adopt, or whether there are barriers that stop that sorting from happening. See \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} for more information on the assumptions involved for the Group Random Coefficients estimator. In our particular case, we contribute to the li

We estimate comparative advantage using the restricted model outlined in \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}:

where $y$ is log yields,

@@

\subsection{Endogenous Switching regression}
The ESS4 is a unique data-set that includes DNA fingerprinting of maize seed type. This brings new possibilities in terms of exploring.... sayESS4 no longer part of panel but that still useful,....//
The group random coefficient model used the variation in the adoption trajectories of each observation in the ESS 1-3 panel as a source of information that allows for an unbiased estimation. ESS4 is no longer part of this panel data-set, hence, in order to estimate the effect of the adoption of improved seeds on yields, an alternative estimation method is needed. We propose an endogenous switching regression model which accounts for unobserved heterogeneity in adoption decisions. This method, although it imposes restrictive assumptions regarding the distribution of errors

\subsection{Group Random Coefficients}

% {\small\tabcolsep=3pt  % hold it local
\resizebox{1\textwidth}{!}{
\input{results/tables/grc_YIELD_cropcutdry_tr}
}

\resizebox{1\textwidth}{!}{
\centering
% \input{results/tables/phi_test_merged}
\input{results/tables/phi_test_self}
\input{results/tables/phi_test_cropcut}
}
% \input{results/tables/grc_YIELD_selfr_tr}

\subsection{Endogenous Switching Regression}

We first examine the results of the Endogenous Switching Regression using a self-reported definition of an improved variety. As before, we consider a more loose definition of improved variety, which includes new seeds but also second-generation and recycled (SR1), and a more strict definition, which includes only new varieties (SR2). Table \ref{tab:switch1} shows the results of the Switching regression using both definitions, and using self-reported and cropcut yields. We see that regardless of the definition of improved seed, yields measurement form and types of controls included, the results consistently show that the Average Treatment effect on the Treated (ATT) is negative and the Average Treatment effect on the Untreated (ATU) is positive, meaning that farmers that adopted a self-reported improved seed would've been better off not adopting, and farmers that did not adopt a self-reported improved seed would've been better off adopting.

\input{results/tables/switch1}

To better understand the above puzzling results, we explore similar yield estimations using the different definitions of improved seeds from the DNA fingerprinting. We run similar switching regressions using different thresholds of DNA purity (70, 90 and 95\% purity), for hybrid seeds (compared with open-pollinated varieties), and for Drought-tolerant maize (DTMZ). Table \ref{tab:switch2} shows the results.

Given that all the plots sampled for DNA fingerprinting showed a purity percent of 70\% or higher, we need a counter-factual to compare these purity thresholds with non-adopters. We therefore consider as non-adopters those that self-reported using a traditional variety. We are aware that given the universality of improved varieties with a 70\% threshold or higher and given the high rates of missclassification, the control group for a 70\% threshold is likely to be contaminated with a high share of false-positives. But we also know the results are biased downwards and that the control group for higher shares of purity will be less contaminated, which may provide an interesting comparison.

Table \ref{tab:switch2} shows that the estimations for purity thresholds of 70 and 90\% show similar patterns as before: non-adopters would've been better off adopting and adopters would've been better off not adopting. But when improved seeds with a purity of 95\% or higher are compared with self-reported traditional varieties or lower purity than 95\%, the ATT becomes positive and the ATU is still positive but significantly decreases (and it is even negative when self-reported yields are used). Farmers that adopted a variety with a 95\% purity or greater had significantly greater yields than if they had not adopted, considering that this is even a lower boundary.

We get similar results for DTMZ. Farmers that adopted DTMZ (as tracked by the DNA germplasm) did obtain greater yields than if they had not adopted. Again this result may be biased downwards because of possible false-positives among the control group. The results for the comparison between hybrid and open-pollinated varieties are not significant (when using self-reported yields).

\input{results/tables/switch2}

To further disaggregate the results, we finally explore the source of the improved varieties (CGIAR or exotic) and the year of the seed release. We find positive effects for adopters of CGIAR-sourced varieties when using self-reported yields. The results seem to be contradictory when using cropcut yields, but we should consider these are different samples, as cropcuts were obtained in a subsample of plots. Exotic-sourced varieties also show positive yield effects for adopters; and a negative ATU, meaning non-adopters are also better off not adopting. Newly released-seed varieties do consistently show positive ATT, when using both self-reported and cropcut yields.

\input{results/tables/switch3}

\section{Discussion and Implications}

- Why do we see that increasing purity leads to positive ATE
- may be due to improved seed being more effective, but doesnt tell the whole story
- barrier to purchasing the seed and associated inputs for poorer households, leading to reuse of seeds and being less effective

@@
